Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-09-09 docs/Manuscript/Manuscript.Rmd
continue addressing reviewer comments

@Kolodner1992 distinguished two styles of CBR: problem-solving, which is more procedural and used when objectives are more clearly defined, and interpretative, which provides categorizations and justifications for possible solutions. A similar dichotomy is commonly used to distinguish the performance criterion for the usefulness of predictive models from the interpretability criterion for their usability. In these terms, localized models are highly procedural: The step of fitting a predictive model to a retrieved cohort can be understood as an automated adaptive strategy [@Begum2011], and the parameters that govern cohort retrieval can be tuned alongside model hyperparameters in a ML workflow. Accordingly, localized models have been primarily used for and evaluated on their ability to more accurately predict classes or outcomes. Nevertheless, as some more recent applications have begun to show, they can be used to draw individualized inferences about the importance of different risk factors. While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected by the user. Provided the model family fitted to the cohorts is interpretable, the localized model as a whole inherits this property. Thus, localized models may embody both styles of CBR.

@@

Our methods followed prepared plans, which we describe in detail in the Appendix.
We also note deviations from these plans and the reasons for them.

@@

**Discuss public availability of code or implementations.**

@@

These programs used varying terminology for common concepts, and no common term is in use for what we here refer to as localized modeling; authors described their approaches as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006c], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015a, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020a], and "precision cohort" analysis [@Wang2019; @Tang2021c; @Ng2021b]. We find uses of "targeted", "personalized", and "precision" generic and imprecise to this approach, while transductive inference is an established term for a broader set of methods in ML. Because _local_ naturally contrasts with _global_, we propose "localized models" as a suitable term for this counterpart to globally-fitted models.[^federated]

[^federated]: As one reviewer pointed out, the terminology of local and global is also used in the unrelated context of federated ML to distinguish models trained using data stored in a single storage device (local) from their aggregations (global) [@Moshawrab2023; @Brauneck2023]. The prospects for federated architecture to more effectively implement similarity cohort--based models are intriguing, and would require some reconciliation of terms. At present, however, we feel that these research areas are sufficiently disjoint to avoid confusion in practice.

>>>>

@Kolodner1992 distinguished two styles of CBR: problem-solving, which is more procedural and used when objectives are more clearly defined, and interpretative, which provides categorizations and justifications for possible solutions. A similar dichotomy is commonly used to distinguish the performance criterion for the usefulness of predictive models from the interpretability criterion for their usability. In these terms, localized models are highly procedural: The step of fitting a predictive model to a retrieved cohort can be understood as an automated adaptive strategy [@Begum2011], and the parameters that govern cohort retrieval can be tuned alongside model hyperparameters in a ML workflow. Accordingly, localized models have been primarily used for and evaluated on their ability to more accurately predict classes or outcomes. Nevertheless, as some more recent applications have begun to show, they can be used to draw inferences about the importance of different risk factors to specific individuals. **Distinguish between interpretable model components like effect estimates and informative but not directly interpretable importance measures as are commonly computed for 'black-box' models.** While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected by the user. Provided the model family fitted to the cohorts is interpretable, the localized model as a whole inherits this property. Thus, localized models may embody both styles of CBR.

@@

We describe the steps of our review in detail in the Appendix.
We also note deviations from plans and the reasons for them.

@@

**Discuss public availability of code or implementations.**
**Methods, results, and inferential reproducibility [@Matarese2022].**

@@

These programs used varying terminology for common concepts, and no common term is in use for what we here refer to as localized modeling; authors described their approaches as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006c], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015a, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020a], and "precision cohort" analysis [@Wang2019; @Tang2021c; @Ng2021b]. We find uses of "targeted", "personalized", and "precision" generic and imprecise to this approach, while transductive inference is an established term for a broader set of methods in ML. Because _local_ naturally contrasts with _global_, we propose "localized models" as a suitable term for this counterpart to globally-fitted models.[^local]

[^local]: We note, in response to one reviewer's comment, that for many if not most models used in ML, excepting GLMs, a predictor may play a different and more or less important role for some cases than others, as quantified by importance measures or model-agnostic explanations [@Biecek2021; @Molnar2023]. In this way, such models can also be viewed as "localized". We recognize that the use of global and local explanations is standard. As another reviewer pointed out, the terminology of local and global is also used in the unrelated context of federated ML to distinguish models trained using data stored in a single storage device (local) from their aggregations (global) [@Moshawrab2023; @Brauneck2023]. The prospects for federated architecture to more effectively implement similarity cohort--based models are intriguing, and would require some reconciliation of terms. At present, we feel that these methods are sufficiently disjoint to avoid confusion in practice, but we suggest the more specific term "similarity cohort--based models" when the term "local" is overloaded.

### 2023-09-09 docs/Manuscript/Manuscript.tex
continue addressing reviewer comments

{[}4{]} distinguished two styles of CBR: problem-solving, which is more
procedural and used when objectives are more clearly defined, and
interpretative, which provides categorizations and justifications for
possible solutions. A similar dichotomy is commonly used to distinguish
the performance criterion for the usefulness of predictive models from
the interpretability criterion for their usability. In these terms,
localized models are highly procedural: The step of fitting a predictive
model to a retrieved cohort can be understood as an automated adaptive
strategy {[}3{]}, and the parameters that govern cohort retrieval can be
tuned alongside model hyperparameters in a ML workflow. Accordingly,
localized models have been primarily used for and evaluated on their
ability to more accurately predict classes or outcomes. Nevertheless, as
some more recent applications have begun to show, they can be used to
draw individualized inferences about the importance of different risk
factors. While the patient similarity measure used to retrieve each
cohort may be complicated, the cohort itself can be directly inspected
by the user. Provided the model family fitted to the cohorts is
interpretable, the localized model as a whole inherits this property.
Thus, localized models may embody both styles of CBR.

@@

\textbf{Discuss public availability of code or implementations.}

@@

These programs used varying terminology for common concepts, and no
common term is in use for what we here refer to as localized modeling;
authors described their approaches as ``targeted prognosis'' {[}11{]},
``transductive inference'' {[}21{]}, ``personalized decision support''
{[}14{]}, ``personalized (predictive) models'' {[}26, in contrast to
``local models,''15,17,32{]}, and ``precision cohort'' analysis
{[}17,34,35{]}. We find uses of ``targeted'', ``personalized'', and
``precision'' generic and imprecise to this approach, while transductive
inference is an established term for a broader set of methods in ML.
Because \emph{local} naturally contrasts with \emph{global}, we propose
``localized models'' as a suitable term for this counterpart to
globally-fitted models.\footnote{As one reviewer pointed out, the
  terminology of local and global is also used in the unrelated context
  of federated ML to distinguish models trained using data stored in a
  single storage device (local) from their aggregations (global)
  {[}39,40{]}. The prospects for federated architecture to more
  effectively implement similarity cohort--based models are intriguing,
  and would require some reconciliation of terms. At present, however,
  we feel that these research areas are sufficiently disjoint to avoid
  confusion in practice.}

@@

The 25 studies included in our synthesis used composite techniques that
we organized into two major types. One type, similarity learning, was
used in 8 studies that tie in to the much larger literature on metric
learning {[}41{]}. These techniques are designed to detect structure in
data, especially associations between predictors and responses, and to
use this structure to inform the definition of a patient similarity
measure. Seven studies used response values in training data sets to
supervise similarity learning while two used unsupervised learning
(Table \ref{tab:composite}). Based on the key properties of metric
learning algorithms identified by {[}42{]}, most learned measures were
non-linear while some were local, and some were optimized globally while
others locally. However, not all qualified as metric learning, since the
resulting measures would not necessarily satisfy the triangle
inequality. Each then used its similarity measure to retrieve training
cases relevant to each testing case. Once retrieved, most took a
standard NN approach to generating predictions from the cohort; two
exceptions {[}29,34{]} proceeded to fit predictive models to the
retrieved cohorts and are discussed below.

@@

The similarity learning approaches took a variety of forms. Several
studies used unsupervised similarity learning, for example the
Mahalanobis distance {[}12{]} and the Kolmogorov entropy-based distance
{[}22{]}. In particular, {[}19{]} defined similarity as a weighted sum
of differences in predictor values and used linear regression to
optimize the weights for the predictive accuracy of the cohort they
retrieve. Others used supervised learning: {[}21{]} proposed an
iterative algorithm to optimize a set of fuzzy inference rules used to
retrieve relevant cases and generate a prediction, using
back-propagation on the rules' parameters. Their method was used in
several later studies returned by our search, which are not discussed
here because they did not originate the technique. {[}28{]} explicitly
appeal to a supervised distance metric learning (DML) technique {[}43{]}
to obtain a similarity measure on their set of dermoscopy and confocal
images that most effectively separates malignant from benign melanoma
tumors. We will discuss the remaining uses of supervised similarity
learning shortly.

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by {[}13{]} to optimize
the predictive performance of models fitted to cohorts retrieved using a
fixed similarity rather than count threshold, a counterpart to
conventional CBR they called ``statistical CBR''.\footnote{Drawing from
  {[}44{]}, we suggest the term ``combinatorial CBR'' for the
  then-conventional approach using similarity cohorts of fixed
  cardinality.} The step used a heuristic procedure to locate a
similarity threshold that (locally) maximizes predictive accuracy on the
training set. This is analogous to optimizing the neighborhood size
parameter in NN predictive modeling, so we will not discuss it further
except to mention that the authors found statistical CBR to outperform
conventional CBR on several data sets. {[}27{]} combined these learning
techniques: They used logistic regression on the training set to obtain
weights for the predictors (by predicting the binary outcome of kidney
transplant waitlist registration) and for the cases (by predicting
agreement of outcome between an index case and other training cases),
and they used exhaustion to optimize the size of the retrieved cohort
for the accuracy of the NN model using the previously optimized weights.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Moshawrab2023}{}}%
\CSLLeftMargin{{[}39{]} }%
\CSLRightInline{Moshawrab M, Adda M, Bouzouane A, Ibrahim H, Raad A.
Reviewing {Federated Machine Learning} and {Its Use} in {Diseases
Prediction}. Sensors 2023;23:2112.
\url{https://doi.org/10.3390/s23042112}.}

\leavevmode\vadjust pre{\hypertarget{ref-Brauneck2023}{}}%
\CSLLeftMargin{{[}40{]} }%
\CSLRightInline{Brauneck A, Schmalhorst L, Majdabadi MMK, Bakhtiari M,
Völker U, Baumbach J, et al. Federated {Machine Learning},
{Privacy-Enhancing Technologies}, and {Data Protection Laws} in {Medical
Research}: {Scoping Review}. Journal of Medical Internet Research
2023;25:e41588. \url{https://doi.org/10.2196/41588}.}

\leavevmode\vadjust pre{\hypertarget{ref-Yang2006}{}}%
\CSLLeftMargin{{[}41{]} }%
\CSLRightInline{Yang L. Distance {Metric Learning}: {A Comprehensive
Survey}. PhD thesis. Michigan State University, 2006.}

\leavevmode\vadjust pre{\hypertarget{ref-Bellet2014}{}}%
\CSLLeftMargin{{[}42{]} }%
\CSLRightInline{Bellet A, Habrard A, Sebban M. A {Survey} on {Metric
Learning} for {Feature Vectors} and {Structured Data} 2014.
\url{https://doi.org/10.48550/arXiv.1306.6709}.}

\leavevmode\vadjust pre{\hypertarget{ref-Xing2002}{}}%
\CSLLeftMargin{{[}43{]} }%
\CSLRightInline{Xing E, Jordan M, Russell SJ, Ng A. Distance {Metric
Learning} with {Application} to {Clustering} with {Side-Information}.
Advances in {Neural Information Processing Systems}, vol. 15, {MIT
Press}; 2002.}

\leavevmode\vadjust pre{\hypertarget{ref-Goyal2008}{}}%
\CSLLeftMargin{{[}44{]} }%
\CSLRightInline{Goyal N, Lifshits Y, Schütze H. Disorder inequality: A
combinatorial approach to nearest neighbor search. Proceedings of the
2008 {International Conference} on {Web Search} and {Data Mining}, {New
York, NY, USA}: {Association for Computing Machinery}; 2008, p. 25--32.
\url{https://doi.org/10.1145/1341531.1341538}.}

>>>>

{[}4{]} distinguished two styles of CBR: problem-solving, which is more
procedural and used when objectives are more clearly defined, and
interpretative, which provides categorizations and justifications for
possible solutions. A similar dichotomy is commonly used to distinguish
the performance criterion for the usefulness of predictive models from
the interpretability criterion for their usability. In these terms,
localized models are highly procedural: The step of fitting a predictive
model to a retrieved cohort can be understood as an automated adaptive
strategy {[}3{]}, and the parameters that govern cohort retrieval can be
tuned alongside model hyperparameters in a ML workflow. Accordingly,
localized models have been primarily used for and evaluated on their
ability to more accurately predict classes or outcomes. Nevertheless, as
some more recent applications have begun to show, they can be used to
draw inferences about the importance of different risk factors to
specific individuals. \textbf{Distinguish between interpretable model
components like effect estimates and informative but not directly
interpretable importance measures as are commonly computed for
`black-box' models.} While the patient similarity measure used to
retrieve each cohort may be complicated, the cohort itself can be
directly inspected by the user. Provided the model family fitted to the
cohorts is interpretable, the localized model as a whole inherits this
property. Thus, localized models may embody both styles of CBR.

@@

\textbf{Discuss public availability of code or implementations.}
\textbf{Methods, results, and inferential reproducibility
{[}\textbf{Matarese2022?}{]}.}

@@

These programs used varying terminology for common concepts, and no
common term is in use for what we here refer to as localized modeling;
authors described their approaches as ``targeted prognosis'' {[}11{]},
``transductive inference'' {[}21{]}, ``personalized decision support''
{[}14{]}, ``personalized (predictive) models'' {[}26, in contrast to
``local models,''15,17,32{]}, and ``precision cohort'' analysis
{[}17,34,35{]}. We find uses of ``targeted'', ``personalized'', and
``precision'' generic and imprecise to this approach, while transductive
inference is an established term for a broader set of methods in ML.
Because \emph{local} naturally contrasts with \emph{global}, we propose
``localized models'' as a suitable term for this counterpart to
globally-fitted models.\footnote{We note, in response to one reviewer's
  comment, that for many if not most models used in ML, excepting GLMs,
  a predictor may play a different and more or less important role for
  some cases than others, as quantified by importance measures or
  model-agnostic explanations {[}39,40{]}. In this way, such models can
  also be viewed as ``localized''. We recognize that the use of global
  and local explanations is standard. As another reviewer pointed out,
  the terminology of local and global is also used in the unrelated
  context of federated ML to distinguish models trained using data
  stored in a single storage device (local) from their aggregations
  (global) {[}41,42{]}. The prospects for federated architecture to more
  effectively implement similarity cohort--based models are intriguing,
  and would require some reconciliation of terms. At present, we feel
  that these methods are sufficiently disjoint to avoid confusion in
  practice, but we suggest the more specific term ``similarity
  cohort--based models'' when the term ``local'' is overloaded.}

@@

The 25 studies included in our synthesis used composite techniques that
we organized into two major types. One type, similarity learning, was
used in 8 studies that tie in to the much larger literature on metric
learning {[}43{]}. These techniques are designed to detect structure in
data, especially associations between predictors and responses, and to
use this structure to inform the definition of a patient similarity
measure. Seven studies used response values in training data sets to
supervise similarity learning while two used unsupervised learning
(Table \ref{tab:composite}). Based on the key properties of metric
learning algorithms identified by {[}44{]}, most learned measures were
non-linear while some were local, and some were optimized globally while
others locally. However, not all qualified as metric learning, since the
resulting measures would not necessarily satisfy the triangle
inequality. Each then used its similarity measure to retrieve training
cases relevant to each testing case. Once retrieved, most took a
standard NN approach to generating predictions from the cohort; two
exceptions {[}29,34{]} proceeded to fit predictive models to the
retrieved cohorts and are discussed below.

@@

The similarity learning approaches took a variety of forms. Several
studies used unsupervised similarity learning, for example the
Mahalanobis distance {[}12{]} and the Kolmogorov entropy-based distance
{[}22{]}. In particular, {[}19{]} defined similarity as a weighted sum
of differences in predictor values and used linear regression to
optimize the weights for the predictive accuracy of the cohort they
retrieve. Others used supervised learning: {[}21{]} proposed an
iterative algorithm to optimize a set of fuzzy inference rules used to
retrieve relevant cases and generate a prediction, using
back-propagation on the rules' parameters. Their method was used in
several later studies returned by our search, which are not discussed
here because they did not originate the technique. {[}28{]} explicitly
appeal to a supervised distance metric learning (DML) technique {[}45{]}
to obtain a similarity measure on their set of dermoscopy and confocal
images that most effectively separates malignant from benign melanoma
tumors. We will discuss the remaining uses of supervised similarity
learning shortly.

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by {[}13{]} to optimize
the predictive performance of models fitted to cohorts retrieved using a
fixed similarity rather than count threshold, a counterpart to
conventional CBR they called ``statistical CBR''.\footnote{Drawing from
  {[}46{]}, we suggest the term ``combinatorial CBR'' for the
  then-conventional approach using similarity cohorts of fixed
  cardinality.} The step used a heuristic procedure to locate a
similarity threshold that (locally) maximizes predictive accuracy on the
training set. This is analogous to optimizing the neighborhood size
parameter in NN predictive modeling, so we will not discuss it further
except to mention that the authors found statistical CBR to outperform
conventional CBR on several data sets. {[}27{]} combined these learning
techniques: They used logistic regression on the training set to obtain
weights for the predictors (by predicting the binary outcome of kidney
transplant waitlist registration) and for the cases (by predicting
agreement of outcome between an index case and other training cases),
and they used exhaustion to optimize the size of the retrieved cohort
for the accuracy of the NN model using the previously optimized weights.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Biecek2021}{}}%
\CSLLeftMargin{{[}39{]} }%
\CSLRightInline{Biecek P, Burzykowski T. Explanatory {Model Analysis}:
{Explore}, {Explain}, and {Examine Predictive Models}. {New York}:
{Chapman and Hall/CRC}; 2021.}

\leavevmode\vadjust pre{\hypertarget{ref-Molnar2023}{}}%
\CSLLeftMargin{{[}40{]} }%
\CSLRightInline{Molnar C. Interpretable {Machine Learning}: {A Guide}
for {Making Black Box Models Explainable}. 2023.}

\leavevmode\vadjust pre{\hypertarget{ref-Moshawrab2023}{}}%
\CSLLeftMargin{{[}41{]} }%
\CSLRightInline{Moshawrab M, Adda M, Bouzouane A, Ibrahim H, Raad A.
Reviewing {Federated Machine Learning} and {Its Use} in {Diseases
Prediction}. Sensors 2023;23:2112.
\url{https://doi.org/10.3390/s23042112}.}

\leavevmode\vadjust pre{\hypertarget{ref-Brauneck2023}{}}%
\CSLLeftMargin{{[}42{]} }%
\CSLRightInline{Brauneck A, Schmalhorst L, Majdabadi MMK, Bakhtiari M,
Völker U, Baumbach J, et al. Federated {Machine Learning},
{Privacy-Enhancing Technologies}, and {Data Protection Laws} in {Medical
Research}: {Scoping Review}. Journal of Medical Internet Research
2023;25:e41588. \url{https://doi.org/10.2196/41588}.}

\leavevmode\vadjust pre{\hypertarget{ref-Yang2006}{}}%
\CSLLeftMargin{{[}43{]} }%
\CSLRightInline{Yang L. Distance {Metric Learning}: {A Comprehensive
Survey}. PhD thesis. Michigan State University, 2006.}

\leavevmode\vadjust pre{\hypertarget{ref-Bellet2014}{}}%
\CSLLeftMargin{{[}44{]} }%
\CSLRightInline{Bellet A, Habrard A, Sebban M. A {Survey} on {Metric
Learning} for {Feature Vectors} and {Structured Data} 2014.
\url{https://doi.org/10.48550/arXiv.1306.6709}.}

\leavevmode\vadjust pre{\hypertarget{ref-Xing2002}{}}%
\CSLLeftMargin{{[}45{]} }%
\CSLRightInline{Xing E, Jordan M, Russell SJ, Ng A. Distance {Metric
Learning} with {Application} to {Clustering} with {Side-Information}.
Advances in {Neural Information Processing Systems}, vol. 15, {MIT
Press}; 2002.}

\leavevmode\vadjust pre{\hypertarget{ref-Goyal2008}{}}%
\CSLLeftMargin{{[}46{]} }%
\CSLRightInline{Goyal N, Lifshits Y, Schütze H. Disorder inequality: A
combinatorial approach to nearest neighbor search. Proceedings of the
2008 {International Conference} on {Web Search} and {Data Mining}, {New
York, NY, USA}: {Association for Computing Machinery}; 2008, p. 25--32.
\url{https://doi.org/10.1145/1341531.1341538}.}
